% !TEX root = ../thesis-index.tex


\chapter*{Abstract}
This thesis investigates the fundamental challenges in training deep neural networks, focusing on signal propagation through network depth. It examines how various architectural choices, such as fully connected layers, weight initialization, normalization layers, and non-linear activations, affect forward and backward passes in deep architectures. The research addresses critical issues such as rank collapse, gradient stability, and their impact on training dynamics and network performance.

Leveraging tools from mean field theory, random matrix theory, and Markov chain theory, we develop a mathematical framework for analyzing signal propagation in deep networks. We characterize conditions leading to rank collapse and gradient instability and provide theoretical insights into the effectiveness of normalization techniques and initialization schemes, suggesting avenues for improving signal propagation and training dynamics in very deep networks. Fundamentally, this thesis's findings are a step toward mathematical principles underlying the success of modern neural network architectures.

\chapter*{Acknowledgments}
I would like to thank my advisor, Prof. Gunnar Ratsch, for his guidance, and mentorship throughout my Ph.D. studies. I would also like to thank my co-advisor, Prof. Francis Bach, for his insightful comments and feedback on my research, which have been instrumental in my academic development. 


During the years, I have had the pleasure of working with and learning from many brilliant researchers. I am particularly thankful for the collaboration with Hadi Daneshmand, who has been a great friend and collaborator throughout my Ph.D. studies. I am also thankful to Dr. Andre Kahles, who supervised and mentored me during the earlier stages of my Ph.D. studies on topics related to bioinformatics. I am thankful to Prof. Thomas Hoffman and Prof. Francesco Orabonoa for their insightful discussions and conversations. 

I would also like to thank former and current members of the biomedical informatics group (BMI) at ETH Zurich, who have provided a warm, supportive, and intellectually stimulating environment. I am grateful to Stefan and Natalia for their friendship and support during the most challenging times. I am also grateful to my friends and colleagues, Gideon, Harun, Alex Immer, Ragnar, Vincent, Victor, Matthias, Natalie, Francesco Locattelo, and the rest of the BMI community for their friendship and various insightful discussions. 

Besides my research network, I am grateful for the friendship and intellectual discussions with several friends. In particular, I am grateful for my conversations with David, Stefan, Alex Meterez, Aran, Hoda, Hadi, Atiye, and Andishe, and Claudia. I am also grateful for my friend, Meysam, who passed away, but left a memory of a great friend with me that will always be cherished.

Last but not least, I am grateful for their ceassless and unconditoinal love and support of my sisters Atefeh and Afagh, my mother Farzaneh, and father Hossein, and my fiancée Alice. I am  There is no overstting that without their love and support, I would not have been able to embark and complete this journey.

\newpage

This thesis is dedicated to my father, who left us shortly before finishing this thesis.